Constraints on Cosmic Rays, Magnetic Fields, and Dark Matter from Gamma-Ray Observations of the Coma Cluster of Galaxies with VERITAS and \Fermi

### 2011-11-14 ms.tex
Minor typo in abstract fixed
Updated the VERITAS acknowledgment blurb

\begin{abstract}
  Observations of radio halos and relics in galaxy clusters indicate efficient electron acceleration
  processes; the same processes are even more efficient for protons. Since protons have a cooling
  time which exceeds the Hubble time in clusters, they can accumulate suggesting that clusters may
  also be sources of very high-energy (VHE; $E>100$ GeV) gamma-ray emission resulting from the
  interactions of highly relativistic cosmic-ray (CR) hadrons with the intracluster gas. Reported
  here are VHE gamma-ray observations of the Coma galaxy cluster with the VERITAS array of imaging
  Cherenkov telescopes, with complementing \Fermi Gamma-ray Space Telescope observations at GeV
  energies. No significant gamma-ray emission from the Coma cluster was detected. Flux upper limits
  at the 99\% confidence level were measured to be on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$
  s$^{-1}$ (VERITAS, $>$220 GeV) and $\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$ (\Fermi, 1-3
  GeV), respectively. For VERITAS (\Fermi), the limits are a factor of 60 (20) higher than
  model-predicted minimum gamma-ray fluxes, which assume the radio-emitting electrons of the Coma
  radio halo to be all secondaries from cosmic-ray interactions. We use the gamma-ray flux upper
  limits to constrain the populations of CRs and magnetic fields and in Coma. Using an analytical
  approach informed by cosmological cluster simulations, the CR-to-thermal pressure is constrained
  to be $< 15\%$ from VERITAS data and $< 2.5\%$ from \Fermi data (averaged within the virial
  radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic CR
  interactions, the observations imply a lower limit on the central magnetic field in Coma of
  $1$--$3\mu$G, depending on how the magnetic field declines with radius and on the gamma-ray
  spectral index. Since these values are well below those inferred by Faraday rotation measurements
  in Coma, this makes the hadronic model a very plausible explanation of the Coma radio
  halo. Finally, a constraint on the thermally averaged product of the total self-annihilation cross
  section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$, is derived
  and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}

@@

\acknowledgments
This research was supported by grants from the U.S. Department of Energy, the U.S. National Science
Foundation, and the Smithsonian Institution, by NSERC in Canada, by Science Foundation Ireland, and
by STFC in the UK. We acknowledge the excellent work of the technical support staff at the FLWO and
the collaborating institutions in the construction and operation of the instrument. C.P. gratefully
acknowledges financial support of the Klaus Tschira Foundation.

>>>>

\begin{abstract}
  Observations of radio halos and relics in galaxy clusters indicate efficient electron acceleration
  processes; the same processes are even more efficient for protons. Since protons have a cooling
  time which exceeds the Hubble time in clusters, they can accumulate suggesting that clusters may
  also be sources of very high-energy (VHE; $E>100$ GeV) gamma-ray emission resulting from the
  interactions of highly relativistic cosmic-ray (CR) hadrons with the intracluster gas. Reported
  here are VHE gamma-ray observations of the Coma galaxy cluster with the VERITAS array of imaging
  Cherenkov telescopes, with complementing \Fermi Gamma-ray Space Telescope observations at GeV
  energies. No significant gamma-ray emission from the Coma cluster was detected. Flux upper limits
  at the 99\% confidence level were measured to be on the order of $2-5\times 10^{-8}$ ph. m$^{-2}$
  s$^{-1}$ (VERITAS, $>$220 GeV) and $\sim 3\times 10^{-6}$ ph. m$^{-2}$ s$^{-1}$ (\Fermi, 1-3
  GeV), respectively. For VERITAS (\Fermi), the limits are a factor of 60 (20) higher than
  model-predicted minimum gamma-ray fluxes, which assume the radio-emitting electrons of the Coma
  radio halo to be all secondaries from cosmic-ray interactions. We use the gamma-ray flux upper
  limits to constrain the populations of CRs and magnetic fields in Coma. Using an analytical
  approach informed by cosmological cluster simulations, the CR-to-thermal pressure is constrained
  to be $< 15\%$ from VERITAS data and $< 2.5\%$ from \Fermi data (averaged within the virial
  radius). Assuming that the radio-emitting electrons of the Coma halo result from hadronic CR
  interactions, the observations imply a lower limit on the central magnetic field in Coma of
  $1$--$3\mu$G, depending on how the magnetic field declines with radius and on the gamma-ray
  spectral index. Since these values are well below those inferred by Faraday rotation measurements
  in Coma, this makes the hadronic model a very plausible explanation of the Coma radio
  halo. Finally, a constraint on the thermally averaged product of the total self-annihilation cross
  section and the relative velocity of the dark-matter particles, $\expval{\sigma v}$, is derived
  and found to be $\lesssim 10^{-21}$ cm$^{3}$ s$^{-1}$ for $m_{\chi}\gtrsim 1000$ GeV/c$^{2}$.
\end{abstract}

@@

\acknowledgments
This research is supported by grants from the U.S. Department of Energy Office of Science, the U.S.
National Science Foundation and the Smithsonian Institution, by NSERC in Canada, by Science
Foundation Ireland (SFI 10/RFP/AST2748) and by STFC in the U.K. We acknowledge the excellent work
of the technical support staff at the Fred Lawrence Whipple Observatory and at the collaborating
institutions in the construction and operation of the instrument.
